\chapter{Future work}
\label{chap:future_work}

The screening narrows the design space to a short list of candidates, and the limitations stated in Section~\ref{sec:limitations} set the direction for the work that should follow. The most immediate item is a corrected-mesh rerun of the full eighteen-case matrix. Seven anchor cases were rerun in this thesis, which left the corrected Pareto status of \texttt{CS\_500\_6} and \texttt{CC\_400\_8} partly open; \texttt{CC\_400\_8} is the single highest-priority rerun, since it sits at the head corner of the uncorrected front.

Beyond the screening itself, three extensions follow from the decoupled scope. A coupled RDT-GV-RPT simulation of the shortlisted candidates would replace the extended straight pipe with the physical runner and test whether the design with the best inflow at \texttt{ConeEnd} is also the best for full-system behaviour. A cavitation assessment would quantify the inner-span sensitivity flagged in Section~\ref{sec:radial-flow-quality}, which the present static-pressure diagnostics leave open. A wet-modal analysis of the candidate vanes, accounting for the added mass of the surrounding water, would close the structural side that the transient hydraulic diagnostics do not cover.
%DEAR HÅVARD 10.06: (new content + humanizer) Replaced the template "this chapter is optional" boilerplate (and the block of empty lines below it) with two concrete future-work paragraphs drawn from Section~\ref{sec:limitations} and the synthesis in Ch.4: corrected rerun of all 18 cases (CC_400_8 first), coupled RDT-GV-RPT simulation, cavitation assessment, wet-modal analysis. No new citations introduced. This lead-in now sits in front of your existing turbine-mode section. VKL voice, British spelling, single-layer hedging.







%NOTATER
\section{Turbine-mode operation with transient coupling}
\label{sec:fw-turbine-mode}

The master agreement identifies turbine-mode operation as a point of interest,
and the present work establishes the guide-vane screening that a turbine-mode
study would build on. A meaningful assessment in the reverse-flow direction was
found to require a transient rotor-stator treatment rather than the steady
frozen-rotor setup used for the screening matrix
(Section~\ref{sec:methods-turbine-mode-ref}). In turbine mode the flow enters
from the draft-tube side and passes through the RDT rotor, so the rotor inflow
is shaped by the bend and cone upstream, and the unsteady rotor-stator
interaction is no longer a small correction to a mean field. The recommended
setup is therefore a transient simulation with a transient rotor-stator
interface at the RDT, initialised from a converged field and run over several
rotor revolutions, following the same spectral-analysis methodology already
used for the pump-mode transient runs (Section~\ref{sec:spec}).

The operating range should be centred on the reverse-rotation condition that
Dokset~\cite{Dokset2026} reports as <PLASSHOLDER: presist hva Dokset fant, f.eks.
"the lowest-head-loss reverse condition for this runner, near 360~rpm">. A
short sweep of rotor speeds around this point would establish how the guide-vane
passage loss and the load distribution between the guide vanes and the RDT
respond to reverse-rotation speed. The same matched constant-chord and
constant-solidity pairs used in the pump-mode screening provide a natural
starting set, so that the chord-strategy contrast can be carried across into
turbine mode.

A full turbine-mode campaign would couple the RDT-GV stage to the RPT runner
with a physical runner-exit inflow rather than the uniform axial inlet used in
the present pump-mode screening, since the runner-exit swirl is part of the
turbine-mode boundary condition. This is a separate study in its own right: it
reintroduces the rotating RPT domain, the runner-passage mesh and the stricter
transient timescale control that Section~\ref{sec:rpt-decoupling-justification}
sets out as the reasons the RPT was decoupled in this thesis. It is the natural
continuation of both the present guide-vane screening and the RPT-side work of
Dokset~\cite{Dokset2026}.
\clearpage % Keep this at the end of the chapter