\chapter*{Disclaimer}
\label{front:discl}
\addcontentsline{toc}{chapter}{Disclaimer}

This master's thesis is a continuation of the author's specialization project carried out during Autumn 2025~\cite{Leonsen2025}. The thesis is an independent academic work, but some material is continued from the project work and appears in Chapters~\ref{chap:introduction}, \ref{chap:litterature} and \ref{chap:theory_method}, where it is identified as such.

I hereby affirm that this master's thesis was composed by myself, and that the work herein is my own except where explicitly stated otherwise in the text. This work has not been submitted for any other degree or professional qualification except as specified, nor has it been published. Where the work of others has been used, whether from a printed source, the internet or any other source, it has been acknowledged and referenced.

Artificial-intelligence tools were used to support the preparation of this thesis. ChatGPT (OpenAI) and Claude (Anthropic) were used for assistance with the MATLAB and general scripting workflow and for troubleshooting in the modelling and CAD workflow. They were also used for idea generation, for structuring of the thesis, and for language refinement and proofreading at the sentence level. After using these tools I reviewed and edited the output, and I take full responsibility for the content of the whole thesis.

%DEAR HÅVARD 10.06: (new content) Skrev disclaimeren etter din instruks: oppgir Claude (Anthropic) og ChatGPT (OpenAI), med bruk = idemyldring, arbeid gjennom oppgaven, strukturering, og spraakvask/korrektur paa setningsniva. Tre ting du boer sjekke: (1) Modellversjoner er ikke spesifisert -- NTNU-malen ber om "AI name, version" saa du kan legge til f.eks. "ChatGPT (GPT-4o/5.1)" og hvilken Claude-modell du brukte hvis sensor vil ha det. (2) AI-verktoyene er navngitt i loepende tekst uten \cite; vil du ha formelle bib-oppfoeringer (\cite{OpenAIChatGPT2024} osv.) maa de legges i bibliography.bib foerst. (3) Jeg droppet de spesifikke emnekodene (TEP4521/TEP4906) fra mal-eksempelet siden jeg ikke kunne verifisere dine -- legg dem inn igjen hvis du vil ha dem.

\clearpage % Keep this at the end of the chapter
